---
title: "R Day2"
author: "LinxinXu"
date: "9/18/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
here::here("data","gapminder5.csv")
gapminder<-read.csv(here::here("data","gapminder5.csv"))
str(gapminder)

```

##Converting factors to strings/characters
#rm(list=ls()): remove all things in the environment 
```{r}
gapminder$country <- as.character(gapminder$country)
gapminder$continent <- as.character(gapminder$continent)
mean(gapminder$lifeExp[gapminder$country=="Afghanistan"])
str(gapminder)
```

##for loop: recover gdp for each country
```{r}
obs<-1:nrow(gapminder)
for (i in obs) {
  gapminder[i,"gdp"]<-gapminder[i,"pop"]*gapminder[i,"gdpPercap"]
}

for (i in obs) {
  gapminder[i,"log_gdpPercap"]<-log(gapminder[i,"gdpPercap"])
  gapminder[i,"log_pop"]<-log(gapminder[i,"pop"])
}

#another way 
gapminder$vec_log_gdpPercap<-log(gapminder$gdpPercap)
all(gapminder$vec_log_gdpPercap==gapminder$log_gdpPercap)
```

##mean life expectancy each year
```{r}
years <- unique(gapminder$year)
for (i in years) {
  mean_le <- mean(gapminder$lifeExp[gapminder$year == i], na.rm = T)
  #not separate strings when paste together
  print(paste0(i, ": ", mean_le))
}

conts <- unique(gapminder$continent)
for (i in conts) {
  mean_le<-mean(gapminder$lifeExp[gapminder$continent==i],na.rm = T)
  print(paste0(i, ": ", mean_le))
}
```


##nested for loop: mean life expectancy for each continent for each year
```{r}
for (i in unique(gapminder$continent)) {
  print(paste0("Continent: ", i))
  for (j in unique(gapminder$year)){
    mean_le<-mean(gapminder$lifeExp[gapminder$year==j&gapminder$continent==i],na.rm=T)
    print(paste0(j, ": ", mean_le))
    
  }
}

years <- unique(gapminder$year)
conts <- unique(gapminder$continent)
for (i in conts) {
  print(paste0("Continent: ", i))
  for (j in years) {
    sd_le <- sd(gapminder$lifeExp[gapminder$continent == i & gapminder$year == j], 
                na.rm = T)
    print(paste0(j, ": ", sd_le))
  }
}
```


##apply family: apply
```{r}
#iterates over rows (1) and cols(2), needs usually a matrix 
vars<-gapminder[,c("lifeExp","pop","gdpPercap")]
apply(vars,2,mean) 

#another way
for (i in vars) {
  print(mean(i))
}
```

##lapply: returns a list, iterate over a values in a vector or list, not rows/cols
```{r}
lapply(gapminder, mean)
```

##sapply: returns a simplified list (i.e., a vector)
```{r}
#mean of lifeExp of each year
years <- unique(gapminder$year)
sapply(years, function(x) mean(gapminder$lifeExp[gapminder$year==x]))
#iterate over first arg
```

##while loop
```{r}
i <-  1952 # define the interator
while (i < 1987) {
  sd_le <- sd(gapminder$lifeExp[gapminder$year == i])
  print(paste0(i, ": ", sd_le)) #do not want any separator use
  i <- i + 5 # increase the iterator by the interval between years
  }

# What is the standard deviation for life expectancy for each year between 1987 and 2002 (inclusive)?

i <-  1987 # define the interator
while (i < 2002) {
  sd_le <- sd(gapminder$lifeExp[gapminder$year == i])
  print(paste0(i, ": ", sd_le)) #do not want any separator use
  i <- i + 1 # increase the iterator by the interval between years
  }
```

##if else condition 
```{r}
set.seed(10)
random_year <- sample(years, 1)

if (random_year > 1977) {
  print(random_year)
}

set.seed(1)
random_year <- sample(years, 1)
if (random_year > 1977) {
  print(paste0(random_year, ": ", 
               mean(gapminder$lifeExp[gapminder$year == random_year])))
  } else { 
    print("sorry, random year is less than 1977")
    }

```


##if else in for loop
```{r}
threshold <- 70

for (i in unique(gapminder$continent)) {
   tmp <- mean(gapminder$lifeExp[gapminder$continent==i])
   
   if (tmp < threshold) {
       print(paste("Mean Life Expectancy in", i, "is less than", threshold))
   } else {
       print(paste("Mean Life Expectancy in", i, "is greater than", threshold))
   }
}


##Write a for loop that reports the mean population for years greater than or equal to 1987. Make sure the loop prints a message if the condition is not met!

for (i in unique(gapminder$year)){
  if (i>=1987){
    print(paste("the mean population for", i, "is", mean(gapminder$pop[gapminder$year==i],na.rm = T)))
  }else{
    print("Sorry, the condition is not met!")
  }
}
```

##writing functions 
```{r}
get_values <-
  function(df, variable = "continent") {
    vals <- unique(df[[variable]])
    print(paste0(variable, ": ", vals))
  }


report_mean_sd <- 
  function(df, variable, country) {
    var <- df[[variable]][df$country == country]
    m_le <- mean(var)
    sd_le <- sd(var)
    cat("Country:", country, 
        "\nMean Life Expectancy:", m_le,
        "\nSD Life Expectancy:", sd_le)
  }
report_mean_sd(gapminder, "lifeExp", "Bulgaria")


report_stats<-function (df, variable, continent){
  var<-df[[variable]][df$continent==continent]
  print(summary(var))
}
report_stats(gapminder, "lifeExp", "Asia")
```


##Functions and loops, together
```{r}
#A log-log model relating life expectancy to GDP
viz_lm <-function(df, dv, iv, year) {
  #dependent and independent variable 
  dat <- df[df[["year"]] == year, ]
  y <- log(dat[[dv]])
  x <- log(dat[[iv]])
  fit <- lm(y ~ x)
  plot(y ~ x, main = year,xlab = iv, ylab = dv)
  lines(x, predict(fit), col = 'blue')
  }
viz_lm(gapminder, "lifeExp", "gdpPercap", 1977)
for (i in years) {
  viz_lm(gapminder, "lifeExp", "gdpPercap", i)
}
```








