---
title: "Day3RExercise_XuLinxin"
author: "LinxinXu"
date: "9/19/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


##Exploring the data
```{r}
here::here("data","generation.csv") #find the home directory and paste things together
generation<-read.csv(here::here("data","generation.csv"),stringsAsFactors = F)

here::here("data","imports.csv") #find the home directory and paste things together
imports<-read.csv(here::here("data","imports.csv"),stringsAsFactors = F)

str(generation)
str(imports)

class(generation$datetime)
class(imports$datetime)

##preserve date and time 
library(lubridate)
generation$datetime <- as_datetime(generation$datetime)
class(generation$datetime)
imports$datetime<-as_datetime(imports$datetime)
class(imports$datetime)
```

##Reshaping data--melt, dcast, recast
```{r}
library(reshape2)
#melt: one obs per row, id.vars can hold several vars using c("id","time") 
long_gen <- melt(generation, id.vars = "datetime",variable.name = "source",
                 value.name = "usage")
head(long_gen)
#order gives index, not actual number
long_gen[order(long_gen$datetime)[1:20], ]

merged_energy <- merge(generation, imports, by = "datetime")
dim(merged_energy)
long_merged_energy <- melt(merged_energy, id.vars = "datetime",variable.name =
                             "source",value.name = "usage")
```


##Dplyr:select,filter,mutate,summarize

##Select: subset variables--one_of(),contains(),starts_with(),ends_with(),matches()
```{r}
library(dplyr)
tmp <- select(merged_energy, biogas, biomass, geothermal, solar)
names(tmp)
tmp <- select(merged_energy, -biogas, -biomass, -geothermal, -solar)
names(tmp)
tmp <- select(merged_energy, contains("hydro"), starts_with("bio"))
names(tmp)
```

##Filter:subset observations based on conditions
```{r}
tmp <- filter(merged_energy, imports > 7000)
head(tmp)
#could use | for or instead of and condition 
tmp <- filter(merged_energy, imports > 7000, natural_gas < 7000)
head(tmp)
```

##Mutate--create new varaibles
```{r}
tmp <- mutate(long_merged_energy, log_usage = log(usage))
head(tmp)
tmp <- mutate(long_merged_energy, log_usage = log(usage), usage2 = usage^2, usage3 
              = usage^3)
head(tmp)
```

##Summarize: reduce multiple observations to a single value (e.g., find the mean)
```{r}
summarize(long_merged_energy, total = sum(usage, na.rm = T))
#While piping, the piped dataframe is not changed!
#To refer to the manipulated dataframe, use .
summarize(long_merged_energy, mean_cons = mean(usage, na.rm = T))

```

#Pipe: shortcut--shift cmd m
```{r}
long_merged_energy %>% 
  filter(source == "geothermal") %>% 
  select(-datetime) %>% 
  mutate(log_usage = log(usage)) %>% 
  summarize(mean_log_usage = mean(log_usage, na.rm = T))

#can add dot after (), doing the same thing 
merged_energy %>% 
  select(-datetime) %>% 
  mutate(total_usage = rowSums(., na.rm = T)) %>% 
  summarize(total_usage = sum(total_usage, na.rm = T))

```

```{r,eval=FALSE}
#Wrong way, bc using "merged_energy
merged_energy %>% 
  select(-datetime) %>% 
  mutate(total_usage = rowSums(merged_energy, na.rm = T)) %>% 
  summarize(total_usage = sum(total_usage, na.rm = T))
```


##Exercise
```{r}
merged_energy %>% select(contains("hydro")) %>% mutate(total_hydro=rowSums(.,na.rm = T)) %>% summarize(mean_usage=mean(total_hydro,na.rm=T))
```

##group by & summarise: perform operations by groups of observations!
```{r}
long_merged_energy %>% 
  group_by(source) %>% 
  summarize(sum_usage = sum(usage, na.rm = T))
```


##Exercise
```{r}
merged_energy %>% select(datetime,contains("hydro"),starts_with("bio")) %>% melt(.,id.vars="datetime",variable.name = "source", value.name = "usage") %>% group_by(source) %>% summarise(mean_usage=mean(usage,na.rm = T))
```



#group by VS. for loop
```{r}
gapminder <- read.csv(here::here("data/gapminder5.csv"))
gapminder %>% 
  group_by(year) %>% 
  summarize(mean_le = mean(lifeExp, na.rm = T),
            sd_lf = sd(lifeExp, na.rm = T))
```






